---
title: "EEB313 Project: Complete cleaned-up code"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

##Setup: required libraries

```{r setup, message=F, echo = F}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(maps)
library(sp)
library(sf)
library(s2)
library(rgdal)
library(mapproj)
library(ggfortify)
library(usethis)
library(ggmap)
library(mapview)
library(arules)
library(stringr)
library(lme4)
library(lmerTest)
library(nlme)
library(MuMIn)
library(report)
library(geosphere)
library(osmdata)
library(arsenal)
```

##Input data

```{r read_data}
morph <- read.csv("Morphology_metadata2.csv")
```

##PCA

```{r PCA grouping}
morph_means <- morph %>% 
  group_by(Species) %>%
  filter(!is.na(Wing) & !is.na(Culmen) & !is.na(Mass)) %>%
  summarize(mean(Wing), mean(Culmen), mean(Mass))
```

```{r PCA and graph}
df <- data.frame(morph_means[,-1])

morph_pc <- prcomp(df, scale = TRUE, center = TRUE, retx = T)
summary(morph_pc)

morph_pc$rotation <- -1*morph_pc$rotation
morph_pc$rotation

autoplot(morph_pc, data = morph_means, colour = 'Species', loadings = TRUE, 
         loadings.label = TRUE) +
labs(title = "Morph Biplot for Hawaii 'Big' Island") +
theme_classic()

plot(morph_pc)
```

##Conversion to lat-long and adding landmarks

```{r spatial_view, message = F}
# Identifies ESPG as 6334.
morph_sf <- st_as_sf(morph, coords = c("UTM_X", "UTM_Y"), crs = 6334)

# Transforms coordinates to standard lat-lon format.
morph_wgs84 <- st_transform(morph_sf, crs = 4326)

# Converts geometry class into separate lat-lon columns.
morph <- morph_wgs84 %>% 
  extract(geometry, c('lon', 'lat'), '\\((.*), (.*)\\)', convert = TRUE) %>% 
  as.data.frame()

# Selects for relavent columns.
morph_spa <- morph %>% 
  select(Species, Date, Sex, Wing, Culmen, Mass, lat, lon)

# Previews the data.
#head(morph_spa)

# Uses ggmap to extract map of Hawaii.
register_google(key = "AIzaSyCAOX4U9l7oFpljVfFnF1U-IW-0ve92QTk")
map_hi <- get_map(location = 'Island of Hawaii', zoom = 9, source = "stamen", 
                  maptype = "toner-lite")

# List of natural and anthropogenic effects that can be analyzed.
effects <- list(Mauna_Loa = c(geocode("Mauna Loa"), Type = "Volcano"),
                Kīlauea = c(geocode("Kīlauea"), Type = "Volcano"),
                Hilo = c(geocode("Hilo"), Type = "Urban"),
                Waimea = c(geocode("Waimea"), Type = "Urban"),
                Kona = c(geocode("Kona"), Type = "Urban"),
                Hualalai = c(geocode("Hualalai"), Type = "Mountain"),
                Mauna_Kea = c(geocode("Mauna Kea Access Rd, Hilo, HI 96720, United States"), 
                              Type = "Mountain"),
                Pohakuloa = c(geocode("Pohakuloa Training Area"), Type = "Military"),
                Army_Reserve = c(geocode("470 W Lanikaula St, Hilo, HI 96720, United States"), 
                                 Type = "Military"))

# Formats list as a dataframe.
effects <- as.data.frame(bind_rows(effects, .id = "Name"))

# Plots the data.

ggmap(map_hi) +
  geom_count(data = morph_spa, aes(x = lon, y = lat), alpha = 0.5) +
  labs(title = "Observations on \nHawaii 'Big' Island", x = "Longitude", y = "Latitude", 
       size = "Number of \nObservations") +
  geom_point(data = effects, aes(x = lon, y = lat, shape = Type), color = "Red")

```

##Distance to landmarks calculation

```{r Distance of points from Landmarks calculation loop}
#extract points lat-long data into a matrix
points <- as.matrix(morph_spa[,7:8])

#names of the distance columns in a list
name_loc <- c("dist_mau_loa","dist_kil","dist_hil","dist_wai","dist_kon",
              "dist_hua","dist_mau_kea","dist_poh","dist_arm_res")

#Distance calculation for loop
for(i in 1:9){
  loc <- as.numeric(c(effects[i,3], effects[i,2]))
  dist_x <-  as.data.frame(spDistsN1(points,loc,longlat=T))
  colnames(dist_x) <- name_loc[i]
  morph_spa <- cbind(morph_spa,dist_x)
}
```

##Distance to shoreline calculation

```{r calculate unique shoreline distances, warning = F}

# Uses the previously created sf_object to identify unique sample sites.
morph_wgs84_distinct <- morph_wgs84 %>% 
  distinct(geometry)



# Creates a bounding box of Hawaii coastline data.
osm_box <- getbb(place_name = "Hawaii") %>%
  opq() %>% 
  add_osm_feature("natural", "coastline") %>% 
  osmdata_sf() 


# Uses the dist2line function in geosphere.
dist_coast <- geosphere::dist2Line(p = st_coordinates(morph_wgs84_distinct), 
                        line = st_coordinates(osm_box$osm_lines)[,1:2])

# Creates a lat lon version of unique sampling sites.
morph_spa_distinct <- morph_spa %>% 
  distinct(lat, lon)

# Combine initial data with distance to coastline.
morph_spa_coast.distance <- cbind(morph_spa_distinct %>% 
                                      rename(y = lat, x = lon), dist_coast) %>%
                                      mutate(kilometers = distance/1000)

# Plot distances
ggmap(map_hi) + 
  geom_point(data = morph_spa_distinct,
             aes(x = lon, y = lat)) +
  geom_segment(data = morph_spa_coast.distance,
               aes(x = x, 
                   y = y, 
                   xend = lon, 
                   yend = lat))
  

```

```{r shoreline distances Binning}
morph_spa_coast.binned <- morph_spa_coast.distance %>% 
  mutate(shr_bin = discretize(kilometers, method="interval", breaks=3, 
                              labels=c("near","middle","far")))
```

```{r Binding with original dataframe, and convert shoreline distance bins to factor}
morph_spa <- left_join(morph_spa, morph_spa_coast.binned %>% 
                         select(y, shr_bin),
                       by = c("lat" = "y"))
morph_spa$shr_bin <- as.character(morph_spa$shr_bin)
```

##Filtering out N/A values

```{r Filter NA}
#Filter out NAs
morph_spa_noNA <- morph_spa %>% 
  filter(!is.na(Wing) & !is.na(Culmen) & !is.na(Mass))
```

##Binning distance to random effects

```{r Bin Random Effects}
#Sort Natural Landmarks (random effect) into categories
morph_spa_bins <- morph_spa_noNA %>% 
  mutate(mau_loa_bins = discretize(dist_mau_loa, method="interval", breaks=3, 
                                   labels=c("near","middle","far"))) %>% 
  mutate(kil_bins = discretize(dist_kil, method="interval", breaks=3, 
                               labels=c("near","middle","far"))) %>% 
  mutate(hua_bins = discretize(dist_hua, method="interval", breaks=3, 
                               labels=c("near","middle","far"))) %>% 
  mutate(mau_kea_bins = discretize(dist_mau_kea, method="interval", breaks=3, 
                                   labels=c("near","middle","far")))
```

##Numericise Year

```{r Numericise Year}
#Extract last 2 digits -- Years as numeric variable
morph_spa_bins <- morph_spa_bins %>% 
  mutate(Year=as.numeric(str_sub(morph_spa_bins$Date,-2,-1)))
```

##Base Linerar Mixed Models

Here are the base linear mixed models, one for each of the morphological traits. Note that all fixed effects, random effects and covariate (year) are included.

```{r Mixed model: Wing}
#Mixed model for Wing
wing_lmer <- lmer(Wing~Year+dist_hil+dist_wai+dist_kon+dist_poh+dist_arm_res+
                    (1|mau_loa_bins)+(1|kil_bins)+(1|hua_bins)+(1|mau_kea_bins)+
                    (1|shr_bin),data=morph_spa_bins,REML=F)
summary(wing_lmer)
```

```{r Mixed model: Culmen}
#Mixed model for Culmen
cul_lmer <- lmer(Culmen~Year+dist_hil+dist_wai+dist_kon+dist_poh+dist_arm_res+
                   (1|mau_loa_bins)+(1|kil_bins)+(1|hua_bins)+(1|mau_kea_bins)+
                   (1|shr_bin),data=morph_spa_bins,REML=F)
summary(cul_lmer)
```

```{r Mixed model: Mass}
#Mixed model for Mass
mas_lmer <- lmer(Mass~Year+dist_hil+dist_wai+dist_kon+dist_poh+dist_arm_res+
                   (1|mau_loa_bins)+(1|kil_bins)+(1|hua_bins)+(1|mau_kea_bins)+
                   (1|shr_bin),data=morph_spa_bins,REML=F)
summary(mas_lmer)
```

##Model Dredging

Model Dredging using MuMIn function dredge()

```{r Wing Dredge, warning=FALSE}
wing_lmer_full <- lmer(Wing~Year+dist_hil+dist_wai+dist_kon+dist_poh+dist_arm_res+
                         (1|mau_loa_bins)+(1|kil_bins)+(1|hua_bins)+(1|mau_kea_bins)+
                         (1|shr_bin),data=morph_spa_bins,REML=F, na.action="na.fail")

wing_dredge <- dredge(wing_lmer_full, rank=AICc)
wing_dredge
```

```{r Culmen Dredge, warning=FALSE}
cul_lmer_full <- lmer(Culmen~Year+dist_hil+dist_wai+dist_kon+dist_poh+dist_arm_res+
                        (1|mau_loa_bins)+(1|kil_bins)+(1|hua_bins)+(1|mau_kea_bins)+
                        (1|shr_bin),data=morph_spa_bins,REML=F, na.action="na.fail")

cul_dredge <- dredge(cul_lmer_full, rank=AICc)
cul_dredge
```


```{r Mass Dredge, warning=FALSE}
mass_lmer_full <- lmer(Mass~Year+dist_hil+dist_wai+dist_kon+dist_poh+dist_arm_res+
                         (1|mau_loa_bins)+(1|kil_bins)+(1|hua_bins)+(1|mau_kea_bins)+
                         (1|shr_bin),data=morph_spa_bins,REML=F, na.action="na.fail")

mass_dredge <- dredge(mass_lmer_full, rank=AICc)
mass_dredge
```

##Model Averaging (after dredge)

```{r Model average - Wing}
top_wing_avg <- model.avg(wing_dredge, subset = delta <=2)
summary(top_wing_avg)
```

```{r Model average - Culmen}
top_cul_avg <- model.avg(cul_dredge, subset = delta <=2)
summary(top_cul_avg)
```

```{r Model average - Mass}
top_mass_avg <- model.avg(mass_dredge, subset = delta <=2)
summary(top_mass_avg)
```

```{r coefficients as table}
coefTable(top_wing_avg,full=T)
coefTable(top_cul_avg,full=T)
coefTable(top_mass_avg,full=T)
```

#Appendix A; exporting cleaned-up Dataset

```{r csv exports}
write.csv(morph_spa_bins, "Cleaned_Dataset_final.csv", row.names = FALSE)
```

## R package Citations

```{r citations}
packs <- c("tidyverse","ggplot2","maps","sp","sf","s2","rgdal","mapproj","ggfortify","usethis",
           "ggmap","mapview","arules","stringr","lme4","lmerTest","nlme","MuMIn","report",
           "geosphere","osmdata","arsenal")
for(i in 1:length(packs)){
  citation(packs[i])
}
```

